Staged Abstract Interpreters

### 2019-09-05 paper/80_related.tex
Finish proofreading

\paragraph{Optimizing Static Analysis Through Specialization} The underlying
idea in this paper is closely related to abstract compilation (AC)
\cite{Boucher:1996:ACN:647473.727587}: removing the interpretative overhead of
traversing the syntax tree by specialization. The residual programs of AC can
be either textual (using quasi-quotations) or closures (using higher-order
functions).  Our approach is more flexible between representations and requires
less engineering efforts to refactor the analyzer (Section \ref{cs_ac}).
\citet{Johnson:2013:OAA:2500365.2500604} adapted closure generation to optimize
small-step abstract interpreter in state-transition style. The program
is first compiled to an "abstract bytecode" IR. The IR consists of higher-order
functions, which will be executed later on an abstract abstract machine for
that IR.  \citet{damian1999partial} provided a formal treatment to abstract
compilation and control-flow analysis, as well as proofs to establish the
correctness of a certified specialized analyzer.  \citet{amtoft1999partial}
applied partial evaluation for constraint-based control-flow analysis.

Splitting an analysis into multiple stages has also been studied for other
forms of static analyses.
\citet{DBLP:conf/cgo/HardekopfL11} applied staging to a
flow-sensitive pointer analysis. The first stage is to analyze the
programs to obtain a sparse representation, and then the second
stage conducts the flow-sensitive analysis by refining the first
one. Another large body of static analyses are implemented by
first generating constraints and then using a solver to discharge the
proof obligations or find the fixed-point. Notable examples include
SAT- and SMT-based program analysis
\cite{Gulwani:2008:PAC:1375581.1375616} and Datalog-based points-to
analysis \cite{Smaragdakis:2015:PA:2802194.2802195}.
Such analyses can be considered having stage distinctions.  One piece of future
work is to explore whether they can be implemented and optimized from
meta-programming. 
Souffl{\'e}, one of the state-of-the-art Datalog engines commonly used in
points-to analysis \cite{Antoniadis:2017:PDS:3088515.3088522}, uses Futamura
projections for efficient implementation \cite{10.1007/978-3-319-41540-6_23}.
However, it only specializes Datalog rules to the running engine, instead of a
full pipeline from a program AST to a specialized solver.

@@

As a source of inspiration of this paper, the Futamura projections reveal a
close relation between interpreters and compilers, which was
originally proposed by \citet{futamura1971partial}, and later reprinted \cite{Futamura1999}.
\citeauthor{Amin:2017:CTI:3177123.3158140} considered a tower of concrete
interpreters and showed how to collapse them by using MSP -- it would be interesting
to explore this idea for multiple abstract interpreters
\cite{Cousot:2019:AAI:3302515.3290355, Giacobazzi:2015:APA:2676726.2676987}.
\citet{10.1007/11561347_18} discussed a monad with let-insertion and
memoization for code generation, using Gaussian Elimination in MetaOCaml as
example. In this work, let-insertion is handled by the ANF transformation in
LMS, and the memoizing monad is just a combination of a reader monad and a state monad
operating on staged data.  Similar to the idea in this paper, specializing
monadic interpreters was explored by \citet{danvy1991compiling,
DBLP:conf/dsl/SheardBP99}. But they are different from the performance motivation
here: the generated code is in monadic form.

>>>>

\paragraph{Optimizing Static Analysis Through Specialization} The underlying
idea in this paper is closely related to abstract compilation (AC)
\cite{Boucher:1996:ACN:647473.727587}: removing the interpretative overhead of
traversing the syntax tree by specialization. The residual programs of AC can
be either textual (using quasi-quotations) or closures (using higher-order
functions).  Our approach is more flexible between representations and requires
less engineering efforts to refactor the analyzer (Section \ref{cs_ac}).
\citet{Johnson:2013:OAA:2500365.2500604} adapted closure generation to optimize
small-step abstract interpreter in state-transition style. The program
is first compiled to an ``abstract bytecode'' IR. The IR consists of higher-order
functions, which will be executed later on an abstract abstract machine for
that IR.  \citet{damian1999partial} provided a formal treatment to abstract
compilation and control-flow analysis, as well as proofs to establish the
correctness of a certified specialized analyzer.  \citet{amtoft1999partial}
applied partial evaluation for constraint-based control-flow analysis.

Splitting an analysis into multiple stages has also been studied for other
forms of static analyses.
\citet{DBLP:conf/cgo/HardekopfL11} applied staging to a
flow-sensitive pointer analysis. The first stage is to analyze the
programs to obtain a sparse representation, and then the second
stage conducts the flow-sensitive analysis by refining the first
one. Another large body of static analyses is implemented by
first generating constraints and then using a solver to discharge the
proof obligations or find the fixed-point. Notable examples include
SAT- and SMT-based program analysis
\cite{Gulwani:2008:PAC:1375581.1375616} and Datalog-based points-to
analysis \cite{Smaragdakis:2015:PA:2802194.2802195}.
Such analyses can be considered having stage distinctions.  One piece of future
work is to explore whether they can be implemented and optimized from
meta-programming.
Souffl{\'e}, one of the state-of-the-art Datalog engines commonly used in
points-to analysis \cite{Antoniadis:2017:PDS:3088515.3088522}, uses Futamura
projections for efficient implementation \cite{10.1007/978-3-319-41540-6_23}.
However, it only specializes Datalog rules to the running engine, instead of a
full pipeline from a program AST to a specialized solver.

@@

As a source of inspiration of this paper, the Futamura projections reveal a
close relation between interpreters and compilers, which was
originally proposed by \citet{futamura1971partial}, and later reprinted \cite{Futamura1999}.
\citeauthor{Amin:2017:CTI:3177123.3158140} considered a tower of concrete
interpreters and showed how to collapse them by using MSP -- it would be interesting
to explore this idea for multiple abstract interpreters
\cite{Cousot:2019:AAI:3302515.3290355, Giacobazzi:2015:APA:2676726.2676987}.
\citet{10.1007/11561347_18} discussed a monad with let-insertion and
memoization for code generation, using Gaussian Elimination in MetaOCaml as an
example. In this work, let-insertion is handled by the ANF transformation in
LMS, and the memoizing monad is just a combination of a reader monad and a state monad
operating on staged data.  Similar to the idea in this paper, specializing
monadic interpreters was explored by \citet{danvy1991compiling,
DBLP:conf/dsl/SheardBP99}. However, they are different from the performance motivation
here: the generated code is in monadic form.

### 2019-09-05 paper/90_conc.tex
typos
Finish proofreading

In this paper, we present a simple and elegant framework that extends the first
Futamura projection to abstract interpreters.
To realize it, we design a generic interpreter that abstracts over both
binding-times and value domains; by instantiating it with staged monads for
abstract interpretation, the staged abstract interpreter can be readily derived.
The meta-programming approach brings new insights and implementation techniques
to optimize static analyzers. Optimizations can be achieved orthogonally to 
the interpreter and abstract semantics.  In our empirical evaluation, staging
improves performance by an order of magnitude in terms of running time.

>>>>

We present a simple and elegant framework that extends the first Futamura
projection to abstract interpreters.
To realize it, we design a generic interpreter that abstracts over both
binding times and value domains; by instantiating it with staged monads for
abstract interpretation, the staged abstract interpreter can be readily derived.
The meta-programming approach brings new insights and implementation techniques
to optimize static analyzers. Optimizations can be achieved orthogonally to
the interpreter and abstract semantics.  In our empirical evaluation, staging
improves performance by an order of magnitude in terms of running time.
